M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2011-12-18 notes/work/ch01.tex
Rename

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Micro-scale phenomena}
Denote positive and negative ionic concentration by $c_+$ and $c_-$ respectively.
The respective fluxes are denoted by $\bj_+$ and $\bj_-$.

@@

The velocity and the pressure field are denoted by $\bv$ and $p$ respectively.

\subsubsection{Nernst-Planck Equations}
The ionic fluxes are given by ions' diffusion, electrostatic forces and ions' advection by the fluid:
\begin{eqnarray}
  \bj_\pm &=& -\bnabla c_+ \mp c_+ \bnabla \varphi + \alpha \bv c_\pm
\end{eqnarray}
Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray}
These equation may be re-written by using the following transform:
\begin{eqnarray}
  c &=& \frac{c_+ + c_-}{2}\\
  q &=& \frac{c_+ - c_-}{2}
\end{eqnarray}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray}
  \bj &=& \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c \\
  \bi &=& \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q
\end{eqnarray}
And are divergence-free as well:
\begin{eqnarray}
\bnabla \cdot \bj &=& 0 \\
\bnabla \cdot \bi &=& 0 
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
Due to strong electrostatic forces, 
most of the fluid remains electroneutral ($c_+ \approx c_-$), 
except of a thin boundary layer (``Debye layer'') around the ion exchanger.

\subsubsection  {Bulk-scale equations}
Nernst-Planck equations above for the fluid bulk (outside
the boundary layer) using bulk variables:
\begin{eqnarray}
  Q & = & 0 \\
  C & = & C_+ = C_- \\
\bJ &=& -\bnabla C + \alpha \bV C \\
\bI &=& -C \bnabla \varPhi
\end{eqnarray}
Since the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:
\begin{eqnarray}
0 = \bnabla \cdot \bJ &\Rightarrow& \Laplacian C - \alpha \bV \cdot \bnabla C = 0 \\
0 = \bnabla \cdot \bI &\Rightarrow& \bnabla \cdot \pars{ C \bnabla \varPhi } = 0
\end{eqnarray}
Stokes flow equations stay the same:
\begin{eqnarray}
\bnabla \cdot \bV &=& 0 \\
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi &=& 0
\end{eqnarray}

@@

In order to match bulk $O(1)$ outer radial ionic flux, 
the $O(\delta^{-1})$ inner ionic radial flux $j_\pm(\rho, \theta)$ must vanish:
\begin{eqnarray}
  -\deriv{c_\pm}{\rho} \mp c_\pm \deriv{\varphi}{\rho} &=& 0 \\
  \deriv{\varphi}{\rho} \pm\frac{1}{c_\pm}\deriv{c_\pm}{\rho} &=& 0 \\
  \pars{\varPhi - \varphi} \pm\pars{\log C_\pm - \log c_\pm} &=& 0 \\
  \varPhi - \varphi \pm \log \frac{C_\pm}{c_\pm} &=& 0
\end{eqnarray}
This relation can be rewritten as Boltzmann distribution of ionic concentration,
since $C_\pm|_{\rho\rightarrow\infty} = C$:
\begin{eqnarray}
c_\pm = C \exp\left[\pm(\varPhi - \varphi)\right]
\end{eqnarray}

@@

Denote zeta potential as the voltage drop between the particle surface and the end of
the boundary layer, $\zeta = \cV - \varPhi$. 
Without loss of generality, 
we assume that $\cV = -\log\gamma$, thus $\zeta = -\log\gamma-\varPhi$:
\begin{eqnarray}
\varPhi &=& -\log C
\end{eqnarray}

Due to anion impermeability and radial flux continuity:
\begin{eqnarray}
  \br \cdot \bj_- = -\deriv{C_-}{r} + C_- \deriv{\varPhi}{r} &=& 0 \\
-\deriv{C}{r} + C \deriv{\varPhi}{r} &=& 0
\end{eqnarray}

The leading-order electric potential in the Debye layer:
\begin{eqnarray}  
\deriv{^2 \varphi}{\rho^2} = -q = -\frac{c_+ - c_-}{2} = 
-C\frac{\exp\left[+(\varPhi - \varphi)\right] - \exp\left[-(\varPhi - \varphi)\right]}{2} = 
C \sinh(\varphi - \varPhi)
\end{eqnarray}
Denote $\psi = \varphi - \varPhi$, such that $\psi(0) = \zeta$ and 
$\psi(\rho\rightarrow\infty) = 0$.
\begin{eqnarray}  
\deriv{^2 \psi}{\rho^2} &=& C \sinh(\psi) = 2 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} 
\\
2\deriv{\psi}{\rho} \deriv{^2 \psi}{\rho^2} &=& 
4 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} \deriv{\psi}{\rho} 
\\
\deriv{}{\rho} \pars{\deriv{\psi}{\rho}}^2 &=& 
\deriv{}{\rho} \pars{2 \sqrt{C} \sinh{\frac{\psi}{2}}}^2 
\end{eqnarray}

Integrating with respect to boundary condtions yields:
\begin{eqnarray}  
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} \\  
\pars{4\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}}^{-1}\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\frac{\pars{\tanh\frac{\psi}{4}}^{-1}}{4\cosh^2\frac{\psi}{4}}
\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\deriv{}{\rho}\pars{\log\tanh\frac{\psi}{4}} &=& \deriv{}{\rho}\pars{-\rho \sqrt{C}} \\
\log\tanh\frac{\psi}{4} - \log\tanh\frac{\zeta}{4} &=& -\rho \sqrt{C} \\
\psi &=& 4 \tanh^{-1}\pars{\tanh{\frac{\zeta}{4}} \cdot \exp\pars{-\rho \sqrt{C}}}
\end{eqnarray}

Leading-order radial momentum balance in the Debye layer yields:
\begin{eqnarray} 
- \deriv{p}{\rho} + \deriv{\varphi}{\rho} \deriv{^2\varphi}{\rho^2} &=& 0
\end{eqnarray}
One integration yields:
\begin{eqnarray} 
p &=& \frac{1}{2}\pars{\deriv{\varphi}{\rho}}^2
\end{eqnarray}

Leading-order tangential momentum balance in the Debye layer yields:
\begin{eqnarray} 
0 &=& \deriv{^2 v_\theta}{\rho^2} - \deriv{p}{\theta} 
 + \deriv{\varphi}{\theta} \deriv{^2\varphi}{\rho^2} 
\\
\deriv{^2 v_\theta}{\rho^2} &=& \frac{1}{2}\deriv{}{\theta}\pars{\deriv{\psi}{\rho}}^2
- \deriv{(\varPhi + \psi)}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{2C \sinh^2\frac{\psi}{2}}
- \deriv{\psi}{\theta} C \sinh\psi
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{C \pars{\cosh\psi - 1}}
- C \sinh\psi \deriv{\psi}{\theta}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{C}{\theta} \pars{\cosh\psi - 1}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = \deriv{C}{\theta} \pars{2 \sinh^2 \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \pars{\half \sinh \frac{\psi}{2} \deriv{\psi}{\rho}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \deriv{}{\rho}\pars{\cosh \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\end{eqnarray}
By integrating from specific $\rho$ to $\rho \rightarrow \infty$:
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} 
\pars{\cosh \frac{\psi}{2} - 1} - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
= -\deriv{C}{\theta} \frac{4}{\sqrt{C}} \sinh^2 \frac{\psi}{4} 
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
\end{eqnarray}
Note that:
\begin{eqnarray}
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} = 
                       -4\sqrt{C}\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}
\\
-4\sinh\frac{\psi}{4} &=& \frac{1}{\sqrt{C} \cosh\frac{\psi}{4}} \deriv{\psi}{\rho}
\end{eqnarray}
Thus:
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& 
  \pars{\deriv{C}{\theta} \frac{1}{C}} \cdot
  \pars{\frac{\sinh\frac{\psi}{4}}{\cosh\frac{\psi}{4}} \deriv{\psi}{\rho}}
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}
=  4\deriv{}{\rho} \pars{\log\cosh\frac{\psi}{4}} \deriv{}{\theta} \log C
  - \deriv{\psi}{\rho} \deriv{\varPhi}{\theta}
\end{eqnarray}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Micro-scale phenomena}

\subsubsection{Dimensionless Notation}
Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$ respectively.

The respective ionic fluxes are denoted by $\bj_+$ and $\bj_-$.

@@

The fluid velocity and the pressure field are denoted by $\bv$ and $p$ respectively.

\subsubsection{Nernst-Planck Equations}
The ionic fluxes are given by ions' diffusion, electrostatic forces and ions' advection by the fluid:
\begin{eqnarray}
  \bj_\pm &=& -\bnabla c_\pm \mp c_\pm \bnabla \varphi + \alpha \bv c_\pm
\end{eqnarray}
Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray}
These equation may be re-written by using the following transform:
\begin{eqnarray}
  c &=& \frac{c_+ + c_-}{2}\\
  q &=& \frac{c_+ - c_-}{2}
\end{eqnarray}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray}
  \bj &=& \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c \\
  \bi &=& \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q
\end{eqnarray}
And are divergence-free as well:
\begin{eqnarray}
\bnabla \cdot \bj &=& 0 \\
\bnabla \cdot \bi &=& 0 
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Separation of scales}
When $\delta \ll 1$, 
most of the fluid remains electroneutral ($c_+ \approx c_-$), 
except of a thin boundary layer of width $O(\delta)$ 
, also known as ``Debye layer'', around the ion exchanger.

\subsubsection  {Bulk-scale equations}
Nernst-Planck equations above for the fluid bulk (outside
the boundary layer) using bulk variables:
\begin{eqnarray}
  Q & = & 0 \\
  C & = & C_+ = C_- \\
\bJ &=& -\bnabla C + \alpha \bV C \\
\bI &=& -C \bnabla \varPhi
\end{eqnarray}
Since the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations.

\begin{enumerate}
\item Salt conservation macroscale equation:
\begin{eqnarray}
\bnabla \cdot \bJ = \Laplacian C - \alpha \bV \cdot \bnabla C = 0 
\end{eqnarray}

\item Charge conservation macroscale equation:
\begin{eqnarray}
\bnabla \cdot \bI = \bnabla \cdot \pars{ C \bnabla \varPhi } = 0
\end{eqnarray}
\end{enumerate}

Stokes flow equations have the same form:
\begin{eqnarray}
\bnabla \cdot \bV &=& 0 \\
\Laplacian \bV - \bnabla P + \bnabla \varPhi \Laplacian \varPhi &=& 0
\end{eqnarray}

@@

In order to match bulk $O(1)$ outer radial ionic flux, 
the $O(\delta^{-1})$ term of the inner ionic radial flux $j_\pm(\rho, \theta)$ must vanish:
\begin{eqnarray}
  -\deriv{c_\pm}{\rho} \mp c_\pm \deriv{\varphi}{\rho} &=& 0 \\
  \deriv{\varphi}{\rho} \pm\frac{1}{c_\pm}\deriv{c_\pm}{\rho} &=& 0 \\
  \pars{\varPhi - \varphi} \pm\pars{\log C_\pm - \log c_\pm} &=& 0 \\
  \varPhi - \varphi \pm \log \frac{C_\pm}{c_\pm} &=& 0
\end{eqnarray}
This relation can be rewritten as Boltzmann distribution of ionic concentration,
since $C_\pm|_{\rho\rightarrow\infty} = C$:
\begin{eqnarray}
c_\pm = C \exp\left[\pm(\varPhi - \varphi)\right]
\end{eqnarray}

@@

Denote zeta potential as the voltage drop between the particle surface and the end of
the boundary layer, $\zeta = \cV - \varPhi$. 
Without loss of generality, 
we assume that $\cV = -\log\gamma$, thus $\zeta = -\log\gamma-\varPhi$:
\begin{eqnarray}
\varPhi + \log C &=& 0
\end{eqnarray}

Due to anion impermeability and radial flux continuity:
\begin{eqnarray}
  \br \cdot \bj_- = -\deriv{c_-}{r} + c_- \deriv{\varPhi}{r} &=& 0 \\
-\deriv{C}{r} + C \deriv{\varPhi}{r} &=& 0 \\
\deriv{}{r}\pars{\varPhi - \log C} &=& 0
\end{eqnarray}

The leading-order electric potential in the Debye layer:
\begin{eqnarray}  
\deriv{^2 \varphi}{\rho^2} = -q = -\frac{c_+ - c_-}{2} = 
-C\frac{\exp\left[+(\varPhi - \varphi)\right] - \exp\left[-(\varPhi - \varphi)\right]}{2} = 
C \sinh(\varphi - \varPhi)
\end{eqnarray}
Denote $\psi = \varphi - \varPhi$, so that $\psi(0) = \cV - \varPhi = \zeta$ and 
$\psi(\rho\rightarrow\infty) = 0$.
\begin{eqnarray}  
\deriv{^2 \psi}{\rho^2} &=& C \sinh(\psi) = 2 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} 
\\
2\deriv{\psi}{\rho} \deriv{^2 \psi}{\rho^2} &=& 
4 C \sinh{\frac{\psi}{2}}\cosh{\frac{\psi}{2}} \deriv{\psi}{\rho} 
\\
\deriv{}{\rho} \pars{\deriv{\psi}{\rho}}^2 &=& 
\deriv{}{\rho} \pars{2 \sqrt{C} \sinh{\frac{\psi}{2}}}^2 
\end{eqnarray}

Integrating with respect to boundary condtions yields:
\begin{eqnarray}  
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} \\  
\pars{4\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}}^{-1}\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\frac{\pars{\tanh\frac{\psi}{4}}^{-1}}{4\cosh^2\frac{\psi}{4}}
\deriv{\psi}{\rho} &=& -\sqrt{C} \\  
\deriv{}{\rho}\pars{\log\tanh\frac{\psi}{4}} &=& \deriv{}{\rho}\pars{-\rho \sqrt{C}} 
\end{eqnarray}

Integrating with respect to boundary condtions yields:
\begin{eqnarray}
\log\tanh\frac{\psi}{4} - \log\tanh\frac{\zeta}{4} &=& -\rho \sqrt{C} \\
\psi = \varphi - \varPhi &=& 
4 \tanh^{-1}\pars{\tanh{\frac{\zeta}{4}} \cdot \exp\pars{-\rho \sqrt{C}}}
\end{eqnarray}

Leading-order radial momentum balance in the Debye layer yields:
\begin{eqnarray} 
- \deriv{p}{\rho} + \deriv{\varphi}{\rho} \deriv{^2\varphi}{\rho^2} &=& 0
\end{eqnarray}
One integration yields:
\begin{eqnarray} 
p &=& \frac{1}{2}\pars{\deriv{\varphi}{\rho}}^2 = \frac{1}{2}\pars{\deriv{\psi}{\rho}}^2
\end{eqnarray}

Leading-order tangential momentum balance in the Debye layer yields:
\begin{eqnarray} 
0 &=& \deriv{^2 v_\theta}{\rho^2} - \deriv{p}{\theta} 
 + \deriv{\varphi}{\theta} \deriv{^2\varphi}{\rho^2} 
\\
\deriv{^2 v_\theta}{\rho^2} &=& \frac{1}{2}\deriv{}{\theta}\pars{\deriv{\psi}{\rho}}^2
- \deriv{(\psi + \varPhi)}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{2C \sinh^2\frac{\psi}{2}}
- \deriv{\psi}{\theta} C \sinh\psi
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{}{\theta}\pars{C \pars{\cosh\psi - 1}}
- C \sinh\psi \deriv{\psi}{\theta}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& \deriv{C}{\theta} \pars{\cosh\psi - 1}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = \deriv{C}{\theta} \pars{2 \sinh^2 \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\\
 &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \pars{\half \sinh \frac{\psi}{2} \deriv{\psi}{\rho}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
 = -\deriv{C}{\theta} \frac{2}{\sqrt{C}} \deriv{}{\rho}\pars{\cosh \frac{\psi}{2}}
- \deriv{\varPhi}{\theta} \deriv{^2\psi}{\rho^2}  
\end{eqnarray}
By integrating from specific $\rho$ to $\rho \rightarrow \infty$ (where $\psi = 0$):
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& -\deriv{C}{\theta} \frac{2}{\sqrt{C}} 
\pars{\cosh \frac{\psi}{2} - 1} - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
= -\deriv{C}{\theta} \frac{4}{\sqrt{C}} \sinh^2 \frac{\psi}{4} 
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}   
\end{eqnarray}
Note that:
\begin{eqnarray}
\deriv{\psi}{\rho} &=& -2\sqrt{C}\sinh\frac{\psi}{2} = 
                       -4\sqrt{C}\sinh\frac{\psi}{4}\cosh\frac{\psi}{4}
\\
-4\sinh\frac{\psi}{4} &=& \frac{1}{\sqrt{C} \cosh\frac{\psi}{4}} \deriv{\psi}{\rho}
\end{eqnarray}
Thus:
\begin{eqnarray}
\deriv{v_\theta}{\rho} &=& 
  \pars{\deriv{C}{\theta} \frac{1}{C}} \cdot
  \pars{\frac{\sinh\frac{\psi}{4}}{\cosh\frac{\psi}{4}} \deriv{\psi}{\rho}}
  - \deriv{\varPhi}{\theta} \deriv{\psi}{\rho}
=  4\deriv{}{\rho} \pars{\log\cosh\frac{\psi}{4}} \deriv{}{\theta} \log C
  - \deriv{\psi}{\rho} \deriv{\varPhi}{\theta}
\\
\deriv{v_\theta}{\rho} &=& 
\deriv{}{\rho} \pars{ 4\log\cosh\frac{\psi}{4} \deriv{}{\theta} \log C
  - \psi \deriv{\varPhi}{\theta} }
\end{eqnarray}
